Topological Equivariant Artist Model for Visualization Library Architecture

### 2021-09-13 paper/main.tex
continuity and topology

\newcommand{\note}[1]{\textcolor{magenta}{#1}} 
\usepackage[nocompress]{cite}
% *** GRAPHICS RELATED PACKAGES ***
\ifCLASSINFOpdf
   \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{./figures/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
   \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{./figures/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
   \DeclareGraphicsExtensions{.eps}
\fi

@@

\section{Related Work}
This work aims to develop a model for describing visualization transformations that can be translated into visualization library architecture. In doing so, we describe how visualization libraries attempt this goal and  build on work that formally describes what properties of data should be preserved. We restrict the properties of data that should be preserved to

@@

\subsection{Continuity}
\begin{figure}[!h]
  \includegraphics[width=\columnwidth]{k_different_types.png}
  \caption{Continuity is how elements in a data set are connected to each other, which is distinct from how the data is structured. The rows in (a) are discrete, therefore they have discrete continuity as illustrated by the discrete dots. The gaussian in (b) is a 1D continuous function, therefore the continuity of the elements of the gaussian can be represented as a line on an interval (0,1). In (c), every element of the globe is connected to its nearest neighbors, which yields a 2D continuous continuity as illustrated by the square.}
\end{figure}

Continuity is

We care about continuity b/c 
\begin{figure}[!h]
  \includegraphics[width=\columnwidth]{whycontinuity.png}
  \caption{Continuity is implicit in choice of visualization rather than explicitely in choice of data container. The line plots in (b) are generated by a 2D table (a). Structurally this table can be identical to the 2D matrix (a) that generates the image in (c).}
\end{figure}

We express continuity using fiber bundles, which are ...
\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}
\begin{equation}
  \label{eq:fiber_bundle}
  \begin{tikzcd}
      \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
  \end{tikzcd}
\end{equation}

@@

\section{Modeling Visualization Stages: Fiber bundles}

%%- brief intro to artist in most simple form $\vartist \dtotal \rightarrow \gtotal$ 
The \textcolor{artist}{Artist $\mathcal{\vartist}$} is a transformation from

\subsection{Data Bundle \dtotal}
\begin{figure}[h!]
  \includegraphics[width=2.5in]{k_qspace.png}
  \caption{this is gonna be replaced w/ more concrete}
\end{figure}
We use topology to model

@@

\paragraph{Structured Data: Section \dsection}
\begin{equation}
  \begin{tikzcd}
      \dfiber \arrow[r, hook] & \dtotal \arrow[d, "\pi"'] \\
                        & \dbase \arrow[u, "\dsection"', bend right]
  \end{tikzcd}
\end{equation}

\paragraph{Structure: Continuity and Equivariant Actions}
\begin{equation}
  \begin{tikzcd}
      \dfiber_i \arrow[d, "m_j"'] \arrow[rd, "m_j\;\circ\; m_k"] & \\
      \dfiber_i \arrow[r, "m_k"']  & \dfiber_i
  \end{tikzcd}
\end{equation}

\subsubsection{Continuity of Data: sheaf \sheaf}
\begin{figure}[h!]
  \includegraphics[width=2.5in]{sheaf.png}
  \caption{}
\end{figure}

\subsection{Graphic Bundle \gtotal} %probably doesn't all need to be subsections, just figure + short description
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{render.png}
\end{figure}
\begin{equation}
  \begin{tikzcd}[ampersand replacement=\&]
      \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
                        \& \gbase \arrow[u, "\gsection"', bend right]
  \end{tikzcd}
\end{equation}
\paragraph{Continuity: Base Space \gbase}
\paragraph{Equivariance: Fiber \gfiber}
\paragraph{Structured Data: Visual \gsection}

\subsection{Artist}
%% flesh out categorical framing of artist, walk through how eq 16 is part of implementing the artists
%% https://github.com/story645/proposal/blob/main/notes/meetings/2021_08_30.md
\begin{equation}
  \vartist: \sheaf(\dtotal) \rightarrow \sheaf(\gtotal)
  \label{eq:math:artist:artist}
\end{equation}

\section{Union of Artists}
\begin{figure}[!h]
\centering
\subfloat[Artists with shared $\mu_i$ renderered correctly]{\includegraphics[width=1.2in]{combined_artist.png}%
\label{fig_first_case}}
%\hfil
\subfloat[Artists without shared $\mu_i$]{\includegraphics[width=1.2in]{exploding_artist.png}%
\label{fig_second_case}}
\caption{Simulation results for the network.}
\label{fig_sim}
\end{figure}

\section{Construction and formal properties of Artists}

@@

\subsection{Data to Representation: \vchannel} %look at what K&S call these stages
\begin{figure}[!h]
  \centering
  \subfloat[Artists with shared $\mu_i$ renderered correctly]{\includegraphics[width=1.2in]{partial_fixed.png}%
  \label{fig_first_case}}
  %\hfil
  \subfloat[Artists without shared $\mu_i$]{\includegraphics[width=1.2in]{partial_invalid.png}%
  \label{fig_second_case}}
  \caption{Simulation results for the network.}
  \label{fig_sim}
  \end{figure}

@@

We enforce the equivariance constraint
\begin{equation}
  \label{eq:math:artist:nu_commute}
\begin{tikzcd}
  \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
  \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
\end{tikzcd}
\end{equation}

\subsubsection{Representation to Visual: \vmark} % look at what K&S call these stages

\begin{figure}[!h]
  \includegraphics[width=2.5in]{diff_type_q.png}
  \caption{rework this as a commutative box w/ the r in E row associated w/ this qhat(k)}
\end{figure}

\section{Case Study}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{path_of_q.png}
    \caption{add in xi!}
  \label{fig:api}
\end{figure}
We implement the \textbf{arrows} in \autoref{fig:api}

\begin{minted}[]{python}

\end{minted}

>>>>

\newcommand{\note}[1]{\textcolor{magenta}{#1}} 
\usepackage[nocompress]{cite}
\usepackage{hyperref} % autoref
% *** GRAPHICS RELATED PACKAGES ***
\ifCLASSINFOpdf
   \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{./figures/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
   \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
   \graphicspath{{./figures/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
   \DeclareGraphicsExtensions{.eps}
\fi

@@

\section{Related Work}
This work aims to develop a model for describing visualization transformations that can serve as guidance for how to architecture a general purpose visualization library. We define a general purpose visualization library as one that provides non domain specific building block components\cite{wongsuphasawatNavigatingWideWorld2021} for building visualizations, for example functions for converting data to color or encoding data as dots. In this section, we describe how visualization libraries attempt this goal and discuss work that formally describes what properties of data should be preserved in a visualization. We restrict the properties of data that should be preserved to

@@

\subsection{Continuity}
\begin{figure}[!h]
  \includegraphics[width=\columnwidth]{k_different_types.png}
  \caption{Continuity is how elements in a data set are connected to each other, which is distinct from how the data is structured. The rows in (a) are discrete, therefore they have discrete continuity as illustrated by the discrete dots. The gaussian in (b) is a 1D continuous function, therefore the continuity of the elements of the gaussian can be represented as a line on an interval (0,1). In (c), every element of the globe is connected to its nearest neighbors, which yields a 2D continuous continuity as illustrated by the square.}
  \label{fig:related-work:continuity}
\end{figure}

Continuity is a representation of how the elements in a dataset are connected to each other. For example, in \autoref{fig:related-work:continuity}, each station record in the table is independent of the others; therefore, the continuity of the table is discrete. The data provided by the gaussian are points sampled along the curve, therefore the continuity of the points on the line is 1D continuous. Every point on the globe is connected to its 6 nearest cardinal neighboring points (NW, N, NE, E, SE, S, SW, W).

\begin{figure}[!h]
  \includegraphics[width=\columnwidth]{whycontinuity.png}
  \caption{Continuity is implicit in choice of visualization rather than explicitly in choice of data container. The line plots in (b) are generated by a 2D table (a). Structurally this table can be identical to the 2D matrix (a) that generates the image in (c).}
  \label{fig:related-work:visual-algorithm}
\end{figure}

Often continuity is expressed in the choice of visual algorithm (visualization type), as explored in taxonomies by Tory and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004} and Chi\cite{chiTaxonomyVisualizationTechniques2000}. For example, in \autoref{fig:related-work:visual-algorithm} the same table can be interpreted as a set of 1D continuous curves when visualized as a collection of line plots or as a 2D surface when visualized as an image.  This means that often there is no way to express data continuity independent of visualization type, meaning most visualization libraries will allow, for example, visualizing discrete data as a line plot or an image. General purpose visualization libraries-such as Matplotlib\cite{hunterMatplotlib2DGraphics2007}, Vtk\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012}, and D3 \cite{bostockDataDrivenDocuments2011}-carry distinct data models as part of the implementation of each visual algorithm. The lack of unified data model means that each plot in a linked\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991} visualization is treated as independent, as are the transforms converting each field in the data to a visual equivalent.

Domain specific libraries can guarantee consistency because they have a single model of the data in their software design, as discussed in Heer and Agarwal \cite{HeerSoftware2006}'s survey of visualization software design patterns. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.

\subsubsection{Fiber Bundles}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fiberbundle.png}
  \caption{A fiber bundle is mathematical construct that allows us to express the relationship between data and continuity. The \textcolor{total}{total} space \dtotal is the topological space in which the data is embedded. The \textcolor{fiber}{fiber} space \dfiber\ is embedded in \dtotal\ and is the set of all possible values that any
  \note{add big rectangle E}}
  \label{fig:related-work:fiber-bundle}
\end{figure}

A model that allows for expressing data continuity in a general flexible way allows us to express, and therefore be faithful to, the multivariate consistency constraints described by Qu and Hullman\cite{hullmanKeeping2018}. A general data model also allows for consistent adaptation to modern data needs, such as complex, metadata rich, distributed, or streaming data. The mathematical theory of fiber bundles provides one such abstraction that can express complicated dimensionality and continuity without being tied to any one data container type, as proposed by Butler, Bryson, and Pendley\cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. In this paper, we build on their work that proposes using topological spaces to represent different properties of data. Here we present a brief summary of topology, for more information see Hatcher\cite{hatcherAlgebraicTopology2002}, Munkres\cite{munkresElementsAlgebraicTopology1984}, and Bradley et. al. \cite{bradleyTopologyCategoricalApproach2020}.

A topological space a topological space $(X, \mathscr{T})$ is a set $X$ with a topology $\mathscr{T}$. Topologies are collections of open sets

such that the empty set and $X$ are in the collection of open sets $\mathscr{T}$, the union of elements in $\mathscr{T}

Specifically, Butler, Bryson, and Pendley suggest that fiber bundles be the basis of an abstract data model. Fiber bundles are a collection $(\dtotal, \dbase, \dfiber, \pi)$ of topological spaces

\begin{LaTeXdescription}
  \item[\textcolor{total}{Total Space} \dtotal]
  \item[\textcolor{fiber}{Fiber Space} \dfiber]
  \item[\textcolor{base}{Base Space} \dbase]  
\end{LaTeXdescription}

with a projection map $\pi:\dtotal\rightarrow\dbase$ that connects every point in \dtotal\ to a point in \dbase. 
\begin{equation}
  \label{eq:fiber_bundle}
  \begin{tikzcd}
      \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
  \end{tikzcd}
\end{equation}

As indicated by $\hookrightarrow$, the fiber space \dfiber\ is embedded inside the total space \dtotal\. This is illustrated in \autoref{fig:related-work:fiber-bundle}, wherein the data lives in the manifold \dtotal. Each data point has two dimensions, represented as the square fiber \dfiber. While the data is embedded in \dtotal, it's continuity is that each point lies along a line, represented by the base space \dbase.

\subsection{Sheaf Maps \sheaf}
We can use sheafs to ensure continuity even when the data is broken up. Is the glue rules -> we can haz parallism. 
\begin{equation}
  \vartist: \sheaf(\dtotal) \rightarrow \sheaf(\gtotal)
  \label{eq:math:artist:artist}
\end{equation}

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{sheaf.png}
  \caption{}
\end{figure}

@@

\subsubsection{Category Theory}
In this work, we propose that equivariance constraints can be expressed using category theory. Vickers et. al provide a brief introduction to category theory for visualization practitioners \cite{vickersUnderstandingVisualizationFormal2013}, but their work focuses on data, representation, and evocation, while this paper is aims to provide guidance on how the map from data to representation should be implemented.

\section{Artist} 
%%- brief intro to artist in most simple form $\vartist \dtotal \rightarrow \gtotal$ 
The \textcolor{artist}{Artist $\mathcal{\vartist}$} is a transformation from a 
%% flesh out categorical framing of artist, walk through how eq 16 is part of implementing the artists
%% https://github.com/story645/proposal/blob/main/notes/meetings/2021_08_30.md

%% domain of A is category E, with sheaves on E7   
%% range of A is category H, sheaves on H
%% category bundles, functions are bundles maps, 
%% subset/restriction is a type of bundle map - bundle over subset is bundle over whole thing, which induces map on section which goes other way which is restriction
%% bundle map induces a sheaf map, fibers themselves are categories, map from fiber to fiber that's legit in F
%%% what is the set up? category of bundles, bundle maps - fiber have some structure/any structure/no structure - bundle maps should be functors, A is gonna be a functor, H is a category of bundles + sheaves of bundles, A is a functor
%%% we claim that for visualizations, A decomposes as 
%%% nu bundle map on bundle + functor on fibers, Q takes you to H
%%% K to S is pullback of deform retract, A is functor on imageA in H

\begin{equation}
  \vartist: \dtotal \rightarrow \gtotal
  \label{eq:math:artist:artist}
\end{equation}

\subsection{Data Bundle \dtotal}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fiberbundle.png}
  \caption{

  \note{replace with more concrete}}
\end{figure}
We use topology to model

@@

\subsubsection{Structured Data: Section \dsection}
\begin{equation}
  \begin{tikzcd}
      \dfiber \arrow[r, hook] & \dtotal \arrow[d, "\pi"'] \\
                        & \dbase \arrow[u, "\dsection"', bend right]
  \end{tikzcd}
\end{equation}

\subsubsection{Structure: Continuity and Equivariant Actions}
\begin{equation}
  \begin{tikzcd}
      \dfiber_i \arrow[d, "m_j"'] \arrow[rd, "m_j\;\circ\; m_k"] & \\
      \dfiber_i \arrow[r, "m_k"']  & \dfiber_i
  \end{tikzcd}
\end{equation}

\subsection{Graphic Bundle \gtotal} %probably doesn't all need to be subsections, just figure + short description
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{render.png}
\end{figure}

\begin{equation}
  \begin{tikzcd}[ampersand replacement=\&]
      \gfiber \arrow[r, hook] \& \gtotal \arrow[d, "\pi"'] \\
                        \& \gbase \arrow[u, "\gsection"', bend right]
  \end{tikzcd}
\end{equation}
\paragraph{Continuity: Base Space \gbase}
\paragraph{Equivariance: Fiber \gfiber}
\paragraph{Structured Data: Visual \gsection}

\subsection{Union of Artists}
\label{sec:artist:union}
\begin{figure}[!h]
\centering
\subfloat[Union of H]{\includegraphics[width=.4\columnwidth]{exploding_artist.png}%
\label{fig_first_case}}
%\hfil
\subfloat[Artist with shared/harmonized \vsection]{\includegraphics[width=.4\columnwidth]{combined_artist.png}%
\label{fig_second_case}}
\caption{Simulation results for the network.}
\label{fig_sim}
\end{figure}

\section{Constraints: Construction and formal properties of Artists}
Identify the constraints, category theory makes it easier to spell out these constraints, this map has to be equivariant
ordered data & data on trees, there is a monoid that acts on them & is not clear that if your map acts on them there will be a nice map on ordered
why category theory? 
- F->V->D
-> F->V is 1:1 equivariance
-> V->D is glyph equivariance

What are the formal properties and constraints on xi, nu, and qhat

nu = any function f:E\rightarrow E, equiv V\rightarrow V
f: F_i\rightarrow F_i, v_i: F_i\right P_i,

define category of structure preserving transformations on F
  - valid maps F->F
  - user in setting up fibers, sets up what are valid transforms/morphisms
  - nu is a functor

%xi is a deform retract, continuous, surjective

%Q: \Gamma (V) \Gamma(H) - is defined on functions not on outputs

%Qhat to Q can't break continuity when it goes from Q to Q hat

%% xi is screen data binding,  map of glyph back to ideal platonic mark

%% continuity is in the sheafs -> requiring that we have a sheaf map and that nu induces a sheaf map, nu can also act on individual components

%% nus are a sheaf map, act on indivdual fibers, induce a sheaf map on E to V, 
%% pull sheafs and sections up to intro, Q is also sheaf map on S 
%%% Q is a sheaf map, category of structure preserving maps, Q is a functor at least on the image

@@

\subsection{Data to Visual Encodings: \vchannel} %look at what K&S call these stages
Visual Channel Encodings
\begin{figure}[!h]
  \centering
  \subfloat[Artists with shared $\mu_i$ renderered correctly]{\includegraphics[width=1.2in]{partial_fixed.png}%
  \label{fig_first_case}}
  %\hfil
  \subfloat[Artists without shared $\mu_i$]{\includegraphics[width=1.2in]{partial_invalid.png}%
  \label{fig_second_case}}
  \caption{Simulation results for the network.}
  \label{fig_sim}
  \end{figure}

@@

We enforce the equivariance constraint

\begin{equation}
  \label{eq:math:artist:nu_commute}
\begin{tikzcd}
  \dtotal_i \arrow[r] \arrow[r, "\vchannel_i"] \arrow[d, "m_{\delement}"'] & \vtotal_i \arrow[d, "m_{\velement}"] \\
  \dtotal_i \arrow[r, "\vchannel_i"]                           & \vtotal_i               
\end{tikzcd}
\end{equation}

\subsubsection{Visual to Graphic: \vmark} % look at what K&S call these stages
Visual encodings to something like marks 
\begin{figure}[!h]
  \includegraphics[width=\columnwidth]{diff_type_q.png}
  \caption{rework this as a commutative box w/ the r in E row associated w/ this qhat(k)}
\end{figure}

\section{Case Study}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{path_of_q.png}
    \caption{\note{add in xi!}}
  \label{fig:api}
\end{figure}
We implement the \textbf{arrows} in \autoref{fig:api}. \mintinline{python}{axesArtist} is a parent artist that acts as a screen. This allows for the composition described in \autoref{sec:artist:union}

\subsection{\vartist}
\begin{minted}{python}
for local_tau in axesArtist.artist.data.query(screen_bounds, dpi):
    mu = axesArtist.artist.graphic.mu(local_tau)
    rho = axesArtist.artist.graphic.qhat(**mu)
    H = rho(renderer)
\end{minted}

where the artist is already parameterized with the \vindex\ functions and which fibers they are associated to:

\begin{min

\subsubsection{\vindex}
\subsubsection{\vchannel}
\subsubsection{\vmarkd}
